XRONOS: an open data infrastructure for archaeological chronology

### 2024-12-05 paper.qmd
Add missing references

---
title: "XRONOS: an open data infrastructure for archaeological chronology"
short-title: XRONOS
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    email: joeroe@hey.com
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [writing, editing, curation, software, analysis]
  - name: Clemens Schmid
    orcid: 0000-0003-3448-5715
    affiliation:
    - ref: mpg
    roles: [curation, software, editing]
  - name: Setareh Ebrahimiabareghi
    orcid: 0000-0003-3749-3147
    affiliation:
    - ref: iaw
    roles: [curation, investigation, editing]
  - name: Caroline Heitz
    orcid: 0000-0001-7188-6775
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, editing]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    email: martin.hinz@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, writing, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    url: https://www.iaw.unibe.ch/
  - id: mpg
    name: Max Planck Institute for Geoanthropology
    url: https://www.shh.mpg.de/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world. By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records. It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating. Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it easier to find information of interest and reducing the need for manual data cleaning by end users. In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
keywords: 
  - open data
  - chronology
  - chronometry
  - radiocarbon dating
  - dendrochronology
  - typological dating
#published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
number-sections: true
bibliography: 
  - references.bib
  - data/c14_datasets.bib
format:
  elsevier-pdf:
    keep-tex: true    
    cite-method: citeproc
    journal:
      name: Journal of Computer Applications in Archaeology
      formatting: review
  # Pretty but broken:
  # https://github.com/andrewheiss/hikmah-academic-quarto/issues/7
  # hikmah-pdf:
  #   papersize: a4
  #   mainfont: "Linux Libertine O"
  #   sansfont: "URW Gothic"
  #   biblatex-chicago: true
  #   biblio-style: authordate
  #   biblatexoptions:
  #     # - backend=biber
  #     - autolang=hyphen
  #     - isbn=false
  #     - uniquename=false
#  html:
#    page-layout: article
#    toc: true
#    embed-resources: true
execute:
  echo: false
  cache: true
---

@@

>>>>

title: "XRONOS: an open data infrastructure for archaeological chronology"
short-title: XRONOS
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    email: joeroe@hey.com
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [writing, editing, curation, software, analysis]
  - name: Clemens Schmid
    orcid: 0000-0003-3448-5715
    affiliation:
    - ref: mpg
    roles: [curation, software, editing]
  - name: Setareh Ebrahimiabareghi
    orcid: 0000-0003-3749-3147
    affiliation:
    - ref: iaw
    roles: [curation, investigation, editing]
  - name: Caroline Heitz
    orcid: 0000-0001-7188-6775
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, editing]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    email: martin.hinz@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, writing, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    url: https://www.iaw.unibe.ch/
  - id: mpg
    name: Max Planck Institute for Geoanthropology
    url: https://www.shh.mpg.de/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world. By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records. It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating. Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it easier to find information of interest and reducing the need for manual data cleaning by end users. In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
keywords: 
  - open data
  - chronology
  - chronometry
  - radiocarbon dating
  - dendrochronology
  - typological dating
#published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
number-sections: true
bibliography: 
  - references.bib
  - data/c14_datasets.bib
nocite: | # workaround for lack of table references
  @Al-Abyadh-Balghelam-Dalma-JebelEtAl, @Alcantara2021, @AndersonEtAl2010, @BANADORA, @BarceloAlvarezEtAl2013, @Benz2010, @BirdEtAl2022, @BohnerSchyle2004, @BronkRamseyEtAl2009, @Capriles2023, @CapuzzoEtAl2014, @ClistEtAl2023, @CochraneEtAl2021, @CourtneyMustaphi2016, @CremaEtAl2016, @DeeEtAl2012, @DeeEtAl2013, @dErricoEtAl2011, @deSaulieuEtAl2017, @Diaz-RodriguezEtAl2023, @DouglassEtAl2019, @FlohrEtAl2016, @GajewskiEtAl2011, @GarnettEtAl2023, @GayoEtAl2015, @GillespieEtAl1984, @GregoriodeSouza2020, @GrossmannEtAl2023, @HoebeEtAl2023, @HoggarthEtAl2021, @HuetEtAl2022, @HuetEtAl2024, @IRDD, @KatsianisEtAl2020, @KellyEtAl2022, @KimEtAl2021, @Kudo2018, @KudoEtAl2023, @LinnenluckeEtAl2023, @Lipo2020, @LoftusEtAl2019, @LucariniEtAl2020, @ManningEtAl2016, @Martinez-GrauEtAl2021, @McFadgenEtAl2000, @MichczynskiEtAl1995, @PalmisanoEtAl2017, @PalmisanoEtAl2022, @PalmisanoEtAl2022a, @Pardo-GordoEtAl2023, @Perrin2019, @PetcheyEtAl2022, @Rademaker2024, @RademakerEtAl2013, @RADON, @RADONB, @RADONB2024, @Raetzel-Fabian1999, @RamseyEtAl2010, @ReingruberThissen2005, @ReingruberThissen2009, @ReingruberThissen2017, @Riris2021, @SeidenstickerSchmid2021, @TanindiErdogu2005, @UriarteGonzalezEtAl2017, @VanStrydonckDeRoock, @Vermeersch2024, @WangEtAl2014, @Weninger2022, @WilliamsEtAl2008, @WilliamsEtAl2014, @WilliamsSmith2012, @WilliamsSmith2013
format:
  elsevier-pdf:
    keep-tex: true    
    cite-method: citeproc
    journal:
      name: Journal of Computer Applications in Archaeology
      formatting: review
  # Pretty but broken:
  # https://github.com/andrewheiss/hikmah-academic-quarto/issues/7
  # hikmah-pdf:
  #   papersize: a4
  #   mainfont: "Linux Libertine O"
  #   sansfont: "URW Gothic"
  #   biblatex-chicago: true
  #   biblio-style: authordate
  #   biblatexoptions:
  #     # - backend=biber
  #     - autolang=hyphen
  #     - isbn=false
  #     - uniquename=false
#  html:
#    page-layout: article
#    toc: true
#    embed-resources: true
execute:
  echo: false
  cache: true

@@

# References
